---
title: "Reproducible Research: Peer Assessment 1"
output:  html_document
keep_md: yes
---

# Introduction

This report describes the steps needed to analyze the data of one person:  
number of steps in 5-minutes intervals during 2 months in 2012.  
For a more detailed overview see the Readme.md-file in the repo. 
The data came with the repo, so is in the working directory  
The data can be downloaded from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

# Loading and preprocessing the data

```{r read data, echo=TRUE}

# when data is not dowloaded or unzipped yet:
if(!file.exists("activity.zip")){
        require(downloader)
        require(curl)
        url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
                download.file(url,"activity.zip", method="curl")
}
if (!file.exists("activity.csv")){
        unzip("activity.zip")
}
data<-read.csv("activity.csv",sep=",")
require(lubridate)
data$date<-ymd(data$date)
nrIntervals<-24*60/5 #hours*minutes/interval_length
```

# What is mean total number of steps taken per day?

```{r calculate mean total steps, echo=TRUE}
require(dplyr)
# remove NA
dat<-as.tbl(data)
by_date<- dat %>%
        filter(steps != "NA") %>%
        group_by(date)
totalStepsPerDay<-summarize(by_date,sum(steps))
names(totalStepsPerDay)<-c("date","steps")
meanStepsPerDay=floor(mean(totalStepsPerDay$steps))
medianStepsPerDay=floor(median(totalStepsPerDay$steps))
hist(totalStepsPerDay$steps,xlab="Total steps per day",ylab="Frequency",main="Total steps per day, NA removed")
abline(v=medianStepsPerDay,col="red")
abline(v=meanStepsPerDay, col="green")
text(x=20000,y=20,labels="mean", col="green")
text(x=20000, y=17.5, labels="median", col="red")
```
  
 
The mean number of steps taken per day is `r sprintf("%.0f",meanStepsPerDay)`.  
The median number of steps taken per day is `r sprintf("%.0f",medianStepsPerDay)`.  



## What is the average daily activity pattern?

```{r average daily activity pattern, echo=TRUE}
by_intervals<- dat %>%
        filter(steps != "NA") %>%
        group_by(interval)
averageDailyActivity<-summarize(by_intervals, mean(steps))
names(averageDailyActivity)<-c("interval","steps")
index1<-which(averageDailyActivity$steps==max(averageDailyActivity$steps))
plot(averageDailyActivity$interval, averageDailyActivity$steps, type="l", xlab=" interval", ylab="avg nr steps", main="Average daily steps, NA removed")

```
  

The 5-minute interval on average across all days in the data set that contains the maximum number of steps taken is `r index1[[1]]` of 288, so at 08:35 hrs.


# Imputing missing values  

For determining a strategy for imputing missing variables, it is important to know when these occur.


```{r Imputting missing values strategy, echo=TRUE}
summary(dat$steps) #report the number of NA in steps
summary(dat$date) # report the number of NA in date
summary(dat$interval) # report the number of NA in interval
#find the missing values
index_NA<-is.na(data$steps)
total_NA<-sum(index_NA)
daysNoActivity<-length(unique(dat$date))- length(unique(by_date$date))
intervalsNoActivity<-daysNoActivity*nrIntervals
```

So the number of days with missing data is `r daysNoActivity`, the total number of NA is `r total_NA`, which is 8\*24\*60/5.
It turns out that the missing values are caused by the fact that no data was recorded during 8 days. Taking the average steps per day as an input would give a big shift in the intervals, therefore the missing value is replaced by the average of the interval.
The new data set will be created in the chunk below.

```{r imputting missing values new data set, echo=TRUE }
dataNoNA<-dat
repl<- rep(averageDailyActivity$steps,61)
dataNoNA$steps[index_NA]<-repl[index_NA]
```

The influence of filling is shown in the following histogram

```{r imputting missing values influence of filling NA, echo=TRUE}
by_date_NoNA<- dataNoNA %>%
        group_by(date)
totalStepsPerDayNoNA<-summarize(by_date_NoNA,sum(steps))
names(totalStepsPerDayNoNA)<-c("date","steps")
meanStepsPerDayNoNA=floor(mean(totalStepsPerDayNoNA$steps))
medianStepsPerDayNoNA=floor(median(totalStepsPerDayNoNA$steps))
summary(totalStepsPerDayNoNA)
hist(totalStepsPerDayNoNA$steps,main="Total steps per day, NA filled",
     xlab="Toal steps per day",ylab="Frequency")
abline(v=medianStepsPerDayNoNA,col="red")
abline(v=meanStepsPerDayNoNA, col="green")
text(x=20000,y=20,labels="mean", col="green")
text(x=20000, y=17.5, labels="median", col="red")
```
  
  
The mean of the total number of steps taken per day is now `r sprintf("%.0f",meanStepsPerDayNoNA)`, while the median is `r sprintf("%.0f",medianStepsPerDayNoNA)`.  
The average does not differ from the one of the original data set. The median is now equal to the average.  
There was only 1 step difference between mean and median in the original data set.  
As the NA were always valid for a whole day, the number of days was increased by imputing averaged interval values for NA's.  
The average did not change, as we corrected with the average steps per day (summation of the average steps for each interval). The median will approach the mean more, as we only add values equal to the mean.


# Are there differences in activity patterns between weekdays and weekends?

```{r patterns weekdays vs weekend, echo=TRUE}

require(dplyr)
require(tidyr)
require(ggplot2)
weekdays1<-c("Monday", "Tuesday","Wednesday","Thursday","Friday")
dataNoNA$wDay <-  factor((weekdays(dataNoNA$date) %in% weekdays1),
                         levels=c(FALSE,TRUE), labels=c('weekend', 'weekday'))
dataNoNA$wDay <-as.character(dataNoNA$wDay)
by_intervals_weekday<- dataNoNA %>%
        filter(wDay=="weekday") %>%
        group_by(interval)
by_intervals_weekend<- dataNoNA %>%
        filter(wDay=="weekend") %>%
        group_by(interval)
averageWeekDay<-summarize(by_intervals_weekday, mean(steps))
names(averageWeekDay)<-c("interval","avg nr steps")
averageWeekend<-summarize(by_intervals_weekend, mean(steps))
names(averageWeekend)<-c("interval","avg nr steps")
par(mfrow = c(2, 1))
plot(averageWeekDay,type="l", ylab="avg nr steps", xlab="interval", main="Weekday")
plot(averageWeekend,type="l", ylab="avg nr steps", xlab="interval", main="Weekend")

```

There are differences between weekdays and weekend: in the weekend the number of steps between 10:00 and 16:00 is much higher than on weekdays



